---
title: "EmilyLiu_A6"
author: "Emily Liu"
date: "10/30/2021"
output: html_document
---

```{r}
library(censusapi)
library(tidyverse)
library(tigris)
library(survey)
library(sf)
library(leaflet)
library(mapview)
#knitr::opts_chunk$set(echo = F, warning = F, message = F)

Sys.setenv(CENSUS_KEY="c8aa67e4086b4b5ce3a8717f59faa9a28f611dab")
```


```{r}
# pums_2019_1yr <- getCensus(
#   name = "acs/acs1/pums",
#   vintage = 2019,
#   region = "public use microdata area:*",
#   regionin = "state:06",
#   vars = c(
#     "SERIALNO",
#     "SPORDER",
#     "PWGTP",
#     "WGTP",
#     "YBL",
#     "BLD",
#     "TEN",
#     "MV",
#     "HINCP",
#     "AGEP"
#   )
# )
# saveRDS(pums_2019_1yr, "a6_pums.rds")
pums_2019_1yr <- readRDS("a6_pums.rds")

ca_pumas <-
  pumas("CA", cb = T, progress_bar = F)

sf_boundary <-
  counties("CA", cb = T, progress_bar = F) %>%
  filter(NAME == "San Francisco")

sf_pumas <-
  ca_pumas %>% 
  st_centroid() %>% 
  .[sf_boundary, ] %>% 
  st_drop_geometry() %>% 
  left_join(ca_pumas %>% select(GEOID10)) %>% 
  st_as_sf()

mapview(sf_pumas)

sf_pums <- pums_2019_1yr %>% 
  mutate(
    PUMA = str_pad(public_use_microdata_area,5,"left","0")
  ) %>% 
  filter(PUMA %in% sf_pumas$PUMACE10)
```


```{r}
sf_pums_clean <- sf_pums %>%
  mutate(
    YBL = as.numeric(YBL),
    AGEP = as.numeric(AGEP),
    HINCP = as.numeric(HINCP)
  ) %>% 
  filter(YBL %in% 1:3) %>% #filters for houses before 1960
  arrange(AGEP) %>%
  group_by(SERIALNO) %>%
  summarize_all(first) %>%
  mutate(
    LEADRISK = ifelse(
      AGEP <= 6 & HINCP <= 90000, #adds leadrisk variable for age less than 6 and household income less than 90,000
      1,
      0
   ) 
  )

#changes columns into factors 
sf_pums_factors <- sf_pums_clean %>%
  mutate(across(c(BLD, TEN, MV, PUMA)))

model <- lm(LEADRISK ~ BLD + TEN + MV + PUMA, sf_pums_factors)

logit_model <- glm(
  LEADRISK ~ BLD + TEN + MV + PUMA,
  family = quasibinomial(),
  data = sf_pums_factors
)

summary(logit_model)

exp(coef(logit_model))

sample_predict <- sample_n(sf_pums_factors, 1)
#data.frame(BLD = 6, TEN = 3, MV = 4, PUMA = 07502)

predict(logit_model, sample_predict, type = "response")
```
6.5 assignment
```{r}
sf_pums_predicted <- sf_pums_factors %>%
  mutate(
    PREDICT = predict(logit_model, sf_pums_factors, type = "response")
  )

#generates the prediction for each row
summary_2x2 <-
  sf_pums_predicted %>% 
  mutate(
    LEADRISK = ifelse(
      LEADRISK == 1, 
      "Yes (Leadrisk)", 
      "No (Leadrisk)"
    ), 
    #WGTP in YES column if predict is greater than 10% 
    YES = ifelse(
      PREDICT >= .1,
      as.numeric(WGTP),
      0
    ),
    #WGTP in NO column if predict is less than 10%
    NO = ifelse(
      PREDICT <= 0.1,
      as.numeric(WGTP),
      0
    )) %>% 
  #consolidates YES, NO, YES(LEADRISK) AND NO(LEADRISK) into a 2x2
  group_by(LEADRISK) %>%
  summarize(
      Predicted_yes =
        sum(YES),
      Predicted_no = 
        sum(NO)
    )
```
Households that exceeded the threshold score generally had lower BLD values (mobile or one-family homes), had 3 for the TEN value (renters) and higher MV values (moved into the home longer ago). 

Roughly how many would they mail out based on their threshold score? Total predicted_yes = 6,182

What percentage of postcards do you expect to actually go to low-income households with children? True positive/all positive = 867/6,182 = 14%

What percentage of all at-risk children are likely to be reached? predicted_yes/Total yes(predicted) = 867/(867+4501) = 867/5,368 = 16.2%

To improve my strategy, I could include more variables in the regression to get a more accurate regression and prediction.