---
title: "EmilyLiu_A8"
author: "Emily Liu"
date: "11/14/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tigris)
library(tidyverse)
library(tidycensus)
library(sf)
library(censusapi)
library(leaflet)
library(StatMatch)
```

```{r}
Sys.setenv(CENSUS_KEY="c8aa67e4086b4b5ce3a8717f59faa9a28f611dab")

acs_vars_2019_5yr <-
  listCensusMetadata(
    name = "2019/acs/acs5",
    type = "variables"
  )

bay_county_names <-
  c(
    "Alameda",
    "Contra Costa",
    "Marin",
    "Napa",
    "San Francisco",
    "San Mateo",
    "Santa Clara",
    "Solano",
    "Sonoma"
  )

bay_tracts <-
  tracts("CA", bay_county_names, cb = T, progress_bar = F)

bay_multiple_tract <- 
  getCensus(
    name = "acs/acs5",
    vintage = 2019,
    region = "tract:*",
    regionin = "state:06+county:001,013,041,055,075,081,085,095,097",
    vars = c(
      "B06009_001E",
      "B06009_002E",
      "B06009_003E",
      "B19001_001E",
      "B19001_014E",
      "B19001_015E",
      "B19001_016E",
      "B19001_017E",
      "B19001A_001E"
    )
  ) %>% 
  transmute(
    tract = paste0(state, county, tract),
    perc_college = 1 - (B06009_002E + B06009_003E) / B06009_001E,
    perc_over100k = (B19001_014E + B19001_015E + B19001_016E + B19001_017E) / B19001_001E,
    perc_white = B19001A_001E / B19001_001E
  ) %>% 
  filter(
    !is.na(perc_college), 
    !is.na(perc_over100k),
    !is.na(perc_white)
  )
```


```{r}
obs_matrix <-
  bay_multiple_tract %>% 
  select(
    perc_white, 
    perc_over100k,
    perc_college
  ) %>% 
  as.matrix()

dist_matrix <- mahalanobis.dist(obs_matrix)

rownames(dist_matrix) <- bay_multiple_tract$tract
colnames(dist_matrix) <- bay_multiple_tract$tract

dist_matrix_pairmatch <- dist_matrix
diag(dist_matrix_pairmatch) <- NA

matched_pair_tract <-
  1:nrow(dist_matrix_pairmatch) %>% 
  map_dfr(function(x){
    
    min_index <- which(dist_matrix_pairmatch[x, ] == min(dist_matrix_pairmatch[x, ], na.rm = T))
   
    data.frame(
      tract = bay_multiple_tract$tract[x],
      matched_tract = bay_multiple_tract$tract[min_index]
    )
    
  })
```

```{r}
leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = bay_tracts %>% 
      filter(GEOID == matched_pair_tract[2,1])
  ) %>% 
  addPolygons(
    data = bay_tracts %>% 
      filter(GEOID == matched_pair_tract[2,2])
  ) %>% 
  addPolygons(
    data = bay_tracts %>% 
      filter(GEOID == matched_pair_tract[3,1]),
    color = "green"
  ) %>% 
  addPolygons(
    data = bay_tracts %>% 
      filter(GEOID == matched_pair_tract[3,2]),
    color = "green"
  ) %>% 
  addPolygons(
    data = bay_tracts %>% 
      filter(GEOID == matched_pair_tract[4,1]),
    color = "red"
  ) %>% 
  addPolygons(
    data = bay_tracts %>% 
      filter(GEOID == matched_pair_tract[4,2]),
    color = "red"
  )
```
```{r}
match_set_tract <- dist_matrix["06081611900", ] %>% 
  as.data.frame() %>% 
  rownames_to_column() %>% 
  rename(
    tract = rowname,
    match = "."
  ) %>% 
  right_join(
    bay_multiple_tract
  ) %>% 
  arrange(match) %>% 
  .[1:21, ] %>% 
  left_join(bay_tracts %>% select(tract = GEOID)) %>% 
  st_as_sf()

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = match_set_tract[1, ],
    color = "red"
  ) %>% 
  addPolygons(
    data = match_set_tract[-1, ]
  )
```
```{r}
Sys.setenv(CENSUS_KEY="c8aa67e4086b4b5ce3a8717f59faa9a28f611dab")

ca_pumas <-
  pumas("CA", cb = T, progress_bar = F)

bay_county_names <-
  c(
    "Alameda",
    "Contra Costa",
    "Marin",
    "Napa",
    "San Francisco",
    "San Mateo",
    "Santa Clara",
    "Solano",
    "Sonoma"
  )

bay_counties <-
  counties("CA", cb = T, progress_bar = F) %>%
  filter(NAME %in% bay_county_names)

bay_pumas <-
  ca_pumas %>% 
  st_centroid() %>% 
  .[bay_counties, ] %>% 
  st_drop_geometry() %>% 
  left_join(ca_pumas %>% select(GEOID10)) %>% 
  st_as_sf()
```

```{r}
leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = bay_pumas,
    weight = 1,
    color = "gray",
    label = ~PUMACE10
  ) %>% 
  addMarkers(
    lng = -121.9415017,
    lat = 37.502171
  ) %>% 
  addPolygons(
    data = bay_pumas %>% 
      filter(PUMACE10 == "01309")
  )
```

```{r}
pums_2014_2019 <- readRDS("/Users/emilyliu/Desktop/2021-2022/Future Bay Initiative/GitHub/emilyliu05/pums_2014_2019_wts.rds")

pums_bart <- pums_2014_2019 %>%
  mutate(
    PWGTP = as.numeric(PWGTP),
    bart = ifelse(
      JWTR %in% c("4"),
      PWGTP,
      0
    )
  ) %>% 
  group_by(PUMA, year) %>% 
  summarize(
    pop = sum(PWGTP),
    bart = sum(bart)
  )

pums_pal <- colorNumeric(
  palette = "YlOrRd",
  domain = pums_bart %>% 
    filter(year == 2018) %>% 
    pull(pop)
)

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = pums_bart %>% 
      filter(year == 2018) %>% 
      right_join(bay_pumas %>% select(PUMA = PUMACE10)) %>% 
      st_as_sf(),
    fillColor = ~pums_pal(pop),
    color = "white",
    weight = 1,
    fillOpacity = 0.5,
    label = ~paste0(PUMA,": Population ", pop)
  )
```
start of chapter 6.2 assignment
```{r}
pums_pal <- colorNumeric(
  palette = "GnBu",
  domain = pums_bart %>% 
    filter(year == 2018) %>% 
    pull(bart)
)

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = pums_bart %>% 
      filter(year == 2018) %>% 
      right_join(bay_pumas %>% select(PUMA = PUMACE10)) %>% 
      st_as_sf(),
    fillColor = ~pums_pal(bart),
    color = "white",
    weight = 1,
    fillOpacity = 0.5,
    label = ~paste0(PUMA,": ", bart, " BART commute riders")
  )

pums_bart_clean <-
  pums_bart %>% 
  select(-pop) %>% 
  pivot_wider(
    names_from = year,
    values_from = bart
  )
```
```{r}
obs_matrix <-
  pums_bart_clean %>% 
  ungroup() %>% 
  select(`2014`,`2015`,`2016`,`2017`) %>% 
  as.matrix()

dist_matrix <- mahalanobis.dist(obs_matrix)

rownames(dist_matrix) <- pums_bart_clean$PUMA
colnames(dist_matrix) <- pums_bart_clean$PUMA

match <- dist_matrix["01309",] %>% 
  as.data.frame() %>% 
  rownames_to_column() %>% 
  rename(
    PUMA = rowname,
    match = "."
  ) %>% 
  right_join(
    pums_bart_clean
  ) %>% 
  arrange(match) %>% 
  .[1:11, ] %>% 
  left_join(bay_pumas %>% select(PUMA = PUMACE10)) %>% 
  st_as_sf()

leaflet() %>% 
  addTiles() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = match[1, ],
    color = "red",
    label = ~PUMA
  ) %>% 
  addPolygons(
    data = match[-1, ],
    label = ~PUMA
  )
```

```{r}
match_pumas <-
  match %>% 
  filter(!PUMA %in% c("01309")) %>% 
  st_drop_geometry() %>% 
  select(-match) %>% 
  pivot_longer(
    -PUMA,
    names_to = "year",
    values_to = "bart"
  ) %>%
  group_by(
    year
  ) %>% 
  summarize(
    bart = mean(bart),
    PUMA = "Similar PUMAs"
  )

treatment_pumas <-
  match %>% 
  filter(PUMA %in% c("01309")) %>% 
  select(-match) %>% 
  st_drop_geometry() %>% 
  pivot_longer(
    -PUMA,
    names_to = "year",
    values_to = "bart"
  )

rbind(
  treatment_pumas,
  match_pumas
) %>% 
  ggplot(
    aes(
      x = as.numeric(year),
      y = bart,
      color = PUMA
    )
  ) +
  geom_line() +
  geom_vline(xintercept = 2018, linetype = "dashed") +
  labs(
    title = "Antioch vs. control neighborhoods, BART ridership",
    x = "Year",
    y = "BART commute riders"
  )
```

```{r}
transit_did <- 
match %>% 
  st_drop_geometry() %>% 
  select(-match) %>% 
  pivot_longer(
    -PUMA,
    names_to = "year",
    values_to = "bart"
  ) %>% 
  mutate(
    year = year %>% as.numeric(),
    time = ifelse(year >= 2018, 1, 0),
    treated = ifelse(PUMA == "01309", 1, 0)
  )

did_reg <- lm(bart ~ treated*time, data = transit_did)

summary(did_reg)
```
Results and assumptions

The difference in BART ridership pre-treatment year between Antioch and the similar PUMAs' is 470.68, the average difference in the baseline before 2018 and after 2018 was smaller, at 207.28. Lastly, the difference-in-difference, which means that the Antioch BART station in 2018 had an estimated impact on about 1097 new BART commuters. Since the p-value is less than 0.05, this means that the result is statistically significant and it's likely that this result isn't due to chance.  
Some other considerations of this analysis are that the station opened in mid-2018, and since the data is only available per year and not throughout the year, it's difficult to line up the timing exactly of the responses with the station opening. Another is that other factors weren't controlled for and we only performed a simple regression, which means that variables that could impact train ridership were neglected. Lastly, the replicate weights weren't accounted for, which would allow a single sample to simulate multiple samples and could lead to lower the standard error. 
